Windowed least-squares model reduction for dynamical systems

### 2019-10-07 main.tex
latest

As described in the introduction, although numerical experiments have
demonstrated that LSPG projection often yields more accurate and stable
solutions than Galerkin
projection~\cite{bui_thesis,carlberg_lspg_v_galerkin,carlberg_gnat,carlberg_thesis,parish_apg},
LSPG still suffers from several shortcomings, in particular its complex
dependence on the time discretization, exponentially growing error bounds, and
a lack of optimality for the trajectory defined over the entire time domain.

@@

Space--time projection methods that employ \spaceTimeAcronym\ trial
spaces~\cite{choi_stlspg,constantine_strom,URBAN2012203,Yano2014ASC,benner_st,bui_thesis}
aim to overcome the latter two shortcomings of LSPG projection, which are
shared by all model-reduction methods based on spatial projection followed by
time integration.  Because these methods employ \spaceTimeAcronym\ trial
spaces, they reduce both the spatial and temporal dimensions of the full-order
model; further, they yield error bounds that grow more slowly in time and
their trajectories exhibit an optimality property over the entire time domain.

\spaceTimeAcronym\ trial subspaces approximate the FOM ODE solution
trajectory
	$\stateFOM\in\RR{N}\otimes \timeSpace$ with an approximation that resides in an
	affine space--time trial subspace of dimension $\stdim\ll\fomdim$, i.e., 
	$\approxstate\in \stspaceST$ with $\dim(\stspaceST) =\stdim $, where
\begin{equation}\label{eq:sttrialspace_def}
 \stspaceST \defeq 
	\Range{\stbasis} + 
	\stateFOMIC\otimes\onesFunction
	\subseteq \RR{N} \otimes \timeSpace
\end{equation}
and $\stbasis\in\RR{N \times \stdim}\otimes \timeSpace$ with 
$\stbasis:\timeDummy\mapsto\stbasis(\timeDummy)$ and $\stbasis(0) =
\boldsymbol 0$ 
denotes the space--time trial basis.
Thus, at any time instance $t\in[0,T]$, ROMs that employ the
\spaceTimeAcronym\ trial subspace approximate the state as
\begin{equation}\label{eq:stapprox1}
 \stateFOMArg{}{t} \approx \approxstateArg{}{t}  = \stbasisArg{}{t} \stgenstate + \stateFOMIC,
\end{equation}
where $\stgenstateArg{} \in \RR{\stdim}$ denote the space--time generalized coordinates. 
Critically, comparing the approximations arising from \spatialAcronym\ and
\spaceTimeAcronym\ subspaces in Eqs.~\eqref{eq:affine_trialspace} and \eqref{eq:stapprox1}, respectively,
highlights that the former approximation associates with a time-dependent
generalized coordinates, while the latter approximation associates with a
time-dependent basis matrix.

Substituting the approximation~\eqref{eq:stapprox1} into the FOM
ODE~\eqref{eq:FOM} yields
\begin{equation}\label{eq:st_fom}
\stbasisDotArg{}{t} \stgenstate =  \velocity(\stbasisArg{}{t} \stgenstate +
	\stateFOMIC,t) , \qquad t \in [0,T],
\end{equation}
where 
$\stbasisDot\equiv d\stbasis/d\timeDummy$. We note that the initial conditions
are automatically satisfied from the definition of the \spaceTimeAcronym\
subspace.

Space--time projection methods reduce the number of equations in
\eqref{eq:st_fom} to ensure a unique solution exists.
We now outline one such method: space--time least-squares Petrov--Galerkin (ST-LSPG)
projection~\cite{choi_stlspg}. 
While space--time Galerkin projection \cite{benner_st} is another alternative, it does not
associate with any residual minimization principle, and thus we do not discuss
it further.
\subsubsection{Space--time LSPG projection}  
Analogously to LSPG projection, space--time LSPG projection~\cite{choi_stlspg}
minimizes the (time-discrete) FOM O$\Delta$E residual, but does so
using the \spaceTimeAcronym\ subspace and simultaneously minimizes this
residual over all $N_t$ time instances.
We first introduce the full space--time FOM O$\Delta$E residual for linear
multistep methods as
\begin{align*}
\residST_{} & \vcentcolon (\stateyDiscreteArg{1}, \ldots
	,\stateyDiscreteArg{N_t};\stateFOMIC) \mapsto \begin{bmatrix}
\residLMSArg{1}(\stateyDiscreteArg{1}; \stateFOMIC) 
%\frac{\stateyDiscreteArg{N_t} - \stateyDiscreteArg{N_t-1}  }{\Delta t} -  \velocity(\stateyDiscreteArg{N_t}) \\
\\ 
%\frac{\stateyDiscreteArg{N_t-1} - \stateFOMDiscreteArg{N_t-2}  }{\Delta t} -  \velocity(\stateyDiscreteArg{N_t-1}) \\
\vdots \\
%\frac{\stateyDiscreteArg{1} - \stateFOMIC }{\Delta t} -  \velocity(\stateyDiscreteArg{1}) \\
\residLMSArg{N_t}(\stateyDiscreteArg{N_t};\stateyDiscreteArg{N_t-1},\ldots,\stateyDiscreteArg{N_t - k^{N_t}}) \end{bmatrix}  \\
& \vcentcolon \RR{\fomdim} \otimes \RR{N_t+1} \rightarrow \RR{\fomdim N_t},
\end{align*}
and define the counterpart function acting on space--time generalized
coordinates:
\begin{align*}
\genresidST_{\text{}} & \vcentcolon (\stgenstatey ;\stateFOMIC) \mapsto \begin{bmatrix}
%\frac{\stateyDiscreteArg{N_t} - \stateyDiscreteArg{N_t-1}  }{\Delta t} -  \velocity(\stateyDiscreteArg{N_t}) \\
\residLMSArg{1}\bigg( \stbasisArg{}{t^1}\stgenstatey + \stateIntercept; \stateFOMIC \bigg)
\\ 
\vdots \\
\residLMSArg{N_t}\bigg( \stbasisArg{}{t^{N_t}}\stgenstatey + \stateIntercept; \stbasisArg{}{t^{N_t-1}}\stgenstatey + \stateIntercept ,\ldots,  \stbasisArg{}{t^{N_t-k^{N_t}}}\stgenstatey + \stateIntercept \bigg)  \\ 
\end{bmatrix}  \\
& \vcentcolon \RR{\stdim} \times \RR{N} \rightarrow \RR{\fomdim N_t}. 
\end{align*}
ST-LSPG projection computes the space--time generalized coordinates that
minimize the space--time FOM O$\Delta$E residual:
\begin{equation}\label{eq:stlspg}
	\stgenstate_\text{ST-LSPG} = \underset{ \stgenstatey\in\RR{\stdim} }{\text{arg\,min}}|| \lspgWeightingSTArg{\stgenstatey}  \genresidST_{\text{}}(\stgenstatey;\stateFOMIC) ||_2^2, 
\end{equation}
where $\lspgWeightingSTArg{\cdot} \in \RR{\stdim \times N N_t}$, with $\stdim
\le n_{st} \le N N_t$, is a space--time weighting matrix that can be chosen,
e.g., to enable hyper-reduction.

>>>>

As described in the introduction, although numerical experiments have
demonstrated that LSPG often yields more accurate and stable
solutions than Galerkin~\cite{bui_thesis,carlberg_lspg_v_galerkin,carlberg_gnat,carlberg_thesis,parish_apg},
LSPG still suffers from several shortcomings. In particular, LSPG suffers from its complex
dependence on the time discretization, exponentially growing error bounds, and
lack of optimality for the trajectory defined over the entire time domain.

@@

Space--time projection methods that employ \spaceTimeAcronym\ trial
spaces~\cite{choi_stlspg,constantine_strom,URBAN2012203,Yano2014ASC,benner_st,bui_thesis}
aim to overcome the latter two shortcomings of LSPG, which are
shared by all model-reduction methods based on spatial projection followed by
time integration. \EP{This is misleading, going to delete} Because these methods employ \spaceTimeAcronym\ trial
spaces, they reduce both the spatial and temporal dimensions of the full-order
model; further, they yield error bounds that grow more slowly in time and
their trajectories exhibit an optimality property over the entire time domain.

\spaceTimeAcronym\ trial subspaces approximate the FOM ODE solution
trajectory
	$\stateFOM\in\RR{N}\otimes \timeSpace$ with an approximation that resides in an
	affine space--time trial subspace of dimension $\stdim\ll\fomdim$, i.e., 
	$\approxstate\in \stspaceST$ with $\dim(\stspaceST) =\stdim $, where
\begin{equation}\label{eq:sttrialspace_def}
 \stspaceST \defeq 
	\Range{\stbasis} + 
	\stateFOMIC\otimes\onesFunction
	\subseteq \RR{N} \otimes \timeSpace
\end{equation}
and $\stbasis\in\RR{N \times \stdim}\otimes \timeSpace$ with 
$\stbasis:\timeDummy\mapsto\stbasis(\timeDummy)$ and $\stbasis(0) =
\boldsymbol 0$
denotes the space--time trial basis.\footnote{For simplicity, we consider the case where the trial space is homogeneous on the boundary.} 
Thus, at any time instance $t\in[0,T]$, ROMs that employ the
\spaceTimeAcronym\ trial subspace approximate the state as
\begin{equation}\label{eq:stapprox1}
 \stateFOMArg{}{t} \approx \approxstateArg{}{t}  = \stbasisArg{}{t} \stgenstate + \stateFOMIC,
\end{equation}
where $\stgenstateArg{} \in \RR{\stdim}$ denote the space--time generalized coordinates. 
Critically, comparing the approximations arising from \spatialAcronym\ and
\spaceTimeAcronym\ trial subspaces in Eqs.~\eqref{eq:affine_trialspace} and \eqref{eq:stapprox1}, respectively,
highlights that the former approximation associates with time-dependent
generalized coordinates, while the latter approximation associates with a
time-dependent basis matrix.

Substituting the approximation~\eqref{eq:stapprox1} into the FOM
ODE~\eqref{eq:FOM} yields
\begin{equation}\label{eq:st_fom}
\stbasisDotArg{}{t} \stgenstate =  \velocity(\stbasisArg{}{t} \stgenstate +
	\stateFOMIC,t) , \qquad t \in [0,T],
\end{equation}
where 
$\stbasisDot\equiv d\stbasis/d\timeDummy$. We note that the initial conditions
are automatically satisfied from the definition of the \spaceTimeAcronym\
trial subspace.

Space--time methods reduce the number of equations in
\eqref{eq:st_fom} to ensure a unique solution exists.
We now outline one such method: space--time least-squares Petrov--Galerkin (ST-LSPG)
projection~\cite{choi_stlspg}. 
While space--time Galerkin projection \cite{benner_st} is another alternative, it does not
associate with any residual minimization principle, and thus we do not discuss
it further.
\subsubsection{Space--time LSPG projection}  
Analogously to LSPG projection, space--time LSPG projection~\cite{choi_stlspg}
minimizes the (time-discrete) FOM O$\Delta$E residual, but does so
using the \spaceTimeAcronym\ subspace and simultaneously minimizes this
residual over all $N_t$ time instances.
We first introduce the full space--time FOM O$\Delta$E residual for linear
multistep methods as
\begin{align*}
\residST_{} & \vcentcolon (\stateyDiscreteArg{1}, \ldots
	,\stateyDiscreteArg{N_t};\stateFOMIC) \mapsto \begin{bmatrix}
\residLMSArg{1}(\stateyDiscreteArg{1}; \stateFOMIC) 
%\frac{\stateyDiscreteArg{N_t} - \stateyDiscreteArg{N_t-1}  }{\Delta t} -  \velocity(\stateyDiscreteArg{N_t}) \\
\\ 
%\frac{\stateyDiscreteArg{N_t-1} - \stateFOMDiscreteArg{N_t-2}  }{\Delta t} -  \velocity(\stateyDiscreteArg{N_t-1}) \\
\vdots \\
%\frac{\stateyDiscreteArg{1} - \stateFOMIC }{\Delta t} -  \velocity(\stateyDiscreteArg{1}) \\
\residLMSArg{N_t}(\stateyDiscreteArg{N_t};\stateyDiscreteArg{N_t-1},\ldots,\stateyDiscreteArg{N_t - k^{N_t}}) \end{bmatrix}  \\
& \vcentcolon \RR{\fomdim} \otimes \RR{N_t+1} \rightarrow \RR{\fomdim N_t},
\end{align*}
and define the counterpart function acting on space--time generalized
coordinates:
\begin{align*}
\genresidST_{\text{}} & \vcentcolon (\stgenstatey ;\stateFOMIC) \mapsto \begin{bmatrix}
%\frac{\stateyDiscreteArg{N_t} - \stateyDiscreteArg{N_t-1}  }{\Delta t} -  \velocity(\stateyDiscreteArg{N_t}) \\
\residLMSArg{1}\bigg( \stbasisArg{}{t^1}\stgenstatey + \stateIntercept; \stateFOMIC \bigg)
\\ 
\vdots \\
\residLMSArg{N_t}\bigg( \stbasisArg{}{t^{N_t}}\stgenstatey + \stateIntercept; \stbasisArg{}{t^{N_t-1}}\stgenstatey + \stateIntercept ,\ldots,  \stbasisArg{}{t^{N_t-k^{N_t}}}\stgenstatey + \stateIntercept \bigg)  \\ 
\end{bmatrix}  \\
& \vcentcolon \RR{\stdim} \times \RR{N} \rightarrow \RR{\fomdim N_t}. 
\end{align*}
ST-LSPG projection computes the space--time generalized coordinates that
minimize the space--time FOM O$\Delta$E residual:
\begin{equation}\label{eq:stlspg}
	\stgenstate_\text{ST-LSPG} = \underset{ \stgenstatey\in\RR{\stdim} }{\text{arg\,min}}|| \lspgWeightingSTArg{\stgenstatey}  \genresidST_{\text{}}(\stgenstatey;\stateFOMIC) ||_2^2, 
\end{equation}
where $\lspgWeightingSTArg{\cdot} \in \RR{\stdim \times N N_t}$, with $\stdim
\le n_{st} \le N N_t$, is a space--time weighting matrix that can be chosen,
e.g., to enable hyper-reduction.
